---
title: "Sex Specific Contributions Exploratory Analysis"
output: 
  html_document: 
    keep_md: true
date: "2024-02-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
library(tidyverse)
library(naniar)
library(janitor)
```

## Data

```{r}
dataset <- read_delim("../data/Sex_specific_contribution.csv", delim=";") %>% clean_names()
```


**First, we look at the structure of our dataset and look for null or NA values.**

```{r}
glimpse(dataset)
```

```{r}
summary(dataset)
```

```{r}
miss_var_summary(dataset)
```

There appear to be NAs, but are all accurately represented as NAs. 

## Nest building

**1. We want to understand the distribution of sexes.** 

```{r}
dataset %>% 
  ggplot(aes(x=nest_builder, fill=nest_builder)) + 
  geom_bar(color="black", alpha=0.8) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Distribution of Nest-Builder Sex",
       x="Nest Builder", 
       y="Count") + 
  scale_fill_brewer(palette="YlGnBu")
```

We see that the most common nest builder is female, followed by both parents, then neither, and the least common (which is very uncommon) is a male nest builder. Since the 'neither' category won't tell us much about patterns for nest building, we can filter it from further analysis.

Who are the male nest builders?
```{r}
dataset %>% 
  select(species, nest_builder) %>% 
  filter(nest_builder == "male")
```

**2. Do other factors have correlations with the sex of the nest builder?**

Let's look at how nest site and nest builder correlate.

```{r}
dataset %>%
  filter(nest_builder != "neither") %>% 
  ggplot(aes(x=nest_builder, fill=nest_builder)) + 
  geom_bar(color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Nest-Builder Sex by Site",
       x="Nest Builder", 
       y="Count",
       fill="Nest Builder") + 
  facet_wrap(~nest_site) + 
  scale_fill_brewer(palette="YlGnBu")
```

We see that females tend to take the lead in ground and grass reed nest building. Both parents tend to be more involved in ground holes, ledges, tree bushes, walls, and water, while neither may do work in tree holes. 

Here's another way of visualizing similar info:

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>% 
  ggplot(aes(x=nest_site, fill=nest_builder)) +
  geom_bar(position = "dodge", color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Nest Site by Nest-Builder Sex",
       x="Site", 
       y="Count",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```

We'll also take a look at how nest structure and nest builder correlate.

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>% 
  ggplot(aes(x=nest_builder, fill=nest_builder)) + 
  geom_bar(color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Nest-Builder Sex by Nest Structure",
       x=NULL, 
       y="Count",
       fill="Nest Builder") + 
  facet_wrap(~nest_structure) + 
  scale_fill_brewer(palette="YlGnBu")
```

Alternatively:
```{r}
dataset %>% 
  filter(nest_builder != "neither") %>% 
  ggplot(aes(x=nest_structure, fill=nest_builder)) +
  geom_bar(position = "dodge", color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Nest Structure by Nest-Builder Sex",
       x="Structure", 
       y="Count",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```


What about the correlation between the incubating sex and nest-building sex?

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>% 
  filter(incubating_sex != "NA") %>% 
  ggplot(aes(x=nest_builder, fill=nest_builder)) + 
  geom_bar(color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Nest-Builder Sex by Incubating Sex",
       x=NULL, 
       y="Count",
       fill="Nest Builder") + 
  facet_wrap(~incubating_sex) + 
  scale_fill_brewer(palette="YlGnBu")
```

Same concept, different plot:

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>% 
  filter(incubating_sex != "NA") %>% 
  ggplot(aes(x=incubating_sex, fill=nest_builder)) +
  geom_bar(position = "dodge", color="black", alpha=0.8) +
  theme(axis.text.x = element_text(angle=50, hjust=1),
        plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Incubating Sex by Nest-Builder Sex",
       x="Incubating Sex", 
       y="Count",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```

**3. How does the sex of the nest builder correlate with our continuous variables?**

Nest builder and mean clutch size:

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>%
  filter(clutch_size_mean != "NA") %>% 
  ggplot(aes(x=nest_builder, y=clutch_size_mean, 
             fill=nest_builder)) +
  geom_boxplot(alpha=0.8) +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Mean Clutch Size by Builder Sex",
       x=NULL, 
       y="Mean Clutch Size",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```


Nest builder vs. length breeding season:

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>%
  filter(length_breeding != "NA") %>% 
  ggplot(aes(x=nest_builder, y=length_breeding, 
             fill=nest_builder)) +
  geom_boxplot(alpha=0.8) +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Length of Breeding Season by Builder Sex",
       x=NULL, 
       y="Length of Breeding Season (months)",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```

Nest builder vs. mean breeding latitude:

```{r}
dataset %>% 
  filter(nest_builder != "neither") %>%
  filter(latitude_mean != "NA") %>% 
  ggplot(aes(x=nest_builder, y=latitude_mean, 
             fill=nest_builder)) +
  geom_boxplot(alpha=0.8) +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  labs(title="Breeding Latitude by Builder Sex",
       x=NULL, 
       y="Mean Breeding Latitude",
       fill="Nest Builder") +
  scale_fill_brewer(palette="YlGnBu")
```

## More Exploration

```{r}
na.omit(dataset) %>%
ggplot(aes(x = latitude_mean, y = clutch_size_mean)) +
  geom_density_2d() +
  labs(title="Mean Latitude vs. Mean Clutch Size",
       x="Mean Latitude",
       y="Mean Clutch Size") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r}
na.omit(dataset) %>%
  ggplot(aes(x=latitude_mean, y=clutch_size_mean)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color="red") + 
  labs(x="Mean Latitude", 
       y="Mean Clutch Size", 
       title="Clutch Size vs. Latitude") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r}
na.omit(dataset) %>%
  filter(nest_builder != "neither") %>%
  ggplot(aes(x = latitude_mean, y = clutch_size_mean, color = nest_structure)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(group = nest_structure)) + # Grouping for separate regression lines
  labs(x="Latitude Mean", 
       y="Mean Clutch Size", 
       title="Clutch Size vs. Latitude by Nest Structure") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  scale_color_brewer(palette = "YlGnBu")
```

```{r}
na.omit(dataset) %>%
  ggplot(aes(x=length_breeding, y=latitude_mean)) +
  geom_point() +
  geom_smooth(method="lm", se=F) +
  labs(title="Breeding Season Length Across Latitudes",
       x="Length Breeding Season",
       y="Mean Latitude") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r}
na.omit(dataset) %>%
  ggplot(aes(x = latitude_mean, y = length_breeding)) +
  geom_point() +  
  geom_smooth(method = "loess", color = "blue") +  
  labs(title="Breeding Season Length Across Latitudes",
       x="Mean Latitude",
       y="Length Breeding Season") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

```{r}
na.omit(dataset) %>%
  filter(nest_builder != "neither") %>%
  ggplot( aes(x = nest_structure)) +
  geom_bar(aes(fill = nest_structure), show.legend = FALSE,
           color="black", alpha=0.8) +
  labs(title ="Nest Structure Distribution", 
       x="Nest Structure", 
       y="Count") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10)) +
  scale_fill_brewer(palette="YlGnBu")
```

```{r}
na.omit(dataset) %>%
 ggplot(aes(x = clutch_size_mean)) +
  geom_histogram(aes(fill = incubating_sex), bins = 30, 
                 alpha = 0.8, position = "identity") +
  facet_wrap(~nest_builder) +
  labs(title="Distribution of Clutch Size Mean by Incubating Sex and Nest Builder",
       x = "Clutch Size Mean",
       y = "Frequency",
       fill="Incubating Sex") +
  scale_fill_brewer(palette = "YlGnBu") +
  theme(plot.title = element_text(size=12, face="bold"),
        axis.title.x = element_text(size=10),
        axis.title.y = element_text(size=10))
```

